The Strucplot Framework:\\ Visualizing Multi-way Contingency Tables with \pkg{vcd}

### 2009-02-23 pkg/vcd/paper/Strucplot/JSS.Rnw
rearrange repository
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@690 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\documentclass{Z}

@@

%% almost as usual
\author{David Meyer, Achim Zeileis, \textnormal{and} Kurt 
Hornik\\Wirtschaftsuniversit\"at Wien, Austria}
\title{The Strucplot Framework:\\ Visualizing Multi-way Contingency
  Tables with \pkg{vcd}}

%% for pretty printing and a nice hypersummary also set:
\Plainauthor{David Meyer, Achim Zeileis, Kurt Hornik} %% comma-separated
\Shorttitle{The Strucplot Framework} %% a short title (if necessary)
\Plaintitle{The Strucplot Framework: Visualizing Multi-way Contingency
  Tables with vcd}

%% an abstract and keywords
\Abstract{
This paper has been published in the Journal of Statistical Software
\citep{vcd:Meyer+Zeileis+Hornik:2006b} and 
describes the ``strucplot'' framework for the visualization of
multi-way contingency tables. Strucplot displays include hierarchical
conditional plots such as mosaic, association, and sieve plots, and
can be combined into more complex, specialized plots for visualizing
conditional independence, GLMs, and the results of
independence tests. The framework's modular design allows
flexible customization of the plots' graphical appearance, including 
shading, labeling, spacing, and legend, by means of ``graphical
appearance control'' functions. 
The framework is provided by the \proglang{R} package \pkg{vcd}.
}

@@

\SweaveOpts{engine=R,eps=TRUE,height=6,width=7,results=hide,fig=FALSE,echo=TRUE}
\setkeys{Gin}{width=0.7\textwidth}
%\VignetteIndexEntry{The Strucplot Framework: Visualizing Multi-way Contingency Tables with vcd}
%\VignetteDepends{vcd}
%\VignetteKeywords{contingency tables, mosaic plots, association plots, sieve plots, categorical data, independence, conditional independence, HSV, HCL, residual-based shading, grid, R}
%\VignettePackage{vcd}

@@

\begin{enumerate}
\item Varying the shape of bar plots and mosaic displays to yield, e.g.,
  double-decker plots \citep{vcd:hofmann:2001}, spine plots, or
  spinograms \citep{vcd:hofmann+theus}.
\item Using residual-based shadings to visualize log-linear models
  \citep{vcd:Friendly:1994,vcd:Friendly:2000} and significance of
  statistical tests \citep{vcd:Meyer+Zeileis+Hornik:2003,vcd:Zeileis+Meyer+Hornik:2005}.
\item Using pairs plots and trellis-like layouts for marginal, conditional and
  partial views \citep{vcd:Friendly:1999}.
\item Adding direct user interaction, allowing quick exploration and
  modification of the visualized models 
  \citep{vcd:Unwin+Hawkins+Hofmann:1996,vcd:Theus:2003}.
\item Providing a modular and flexible implementation to easily allow
  user extensions \citep{vcd:Meyer+Zeileis+Hornik:2003,vcd:Meyer+Zeileis+Hornik:2006b}.
\end{enumerate}

\noindent Current implementations of mosaic displays can be found,
e.g., for \proglang{SAS} \citep{vcd:SAS:2005}, \pkg{ViSta} \citep{vcd:young:1996}, 
\pkg{MANET} \citep{vcd:Unwin+Hawkins+Hofmann:1996},
\pkg{Mondrian} \citep{vcd:Theus:2003},
\proglang{R} \citep{vcd:R:2006}, and \proglang{S-PLUS}
\citep{vcd:SPLUS:2005}. For \proglang{R}, currently three
implementations do exist in the packages \pkg{graphics} (in base \proglang{R}), 
\pkg{vcd} \citep{vcd:Meyer+Zeileis+Hornik:2006b},
and \pkg{iplots} \citep{vcd:urbanek+wichtrey:2006}, respectively.
Table \ref{tab:compare} gives an overview of the available functionality in these systems.
Most environments are available on Windows, MacOS, and Linux/Unix
variants, except \pkg{MANET} which is only available for the
Macinthosh platforms.

@@

Figures \ref{fig:arthritis} to \ref{fig:titanic} illustrate some of these
extensions. Figure~\ref{fig:arthritis} shows the results from a
double-blind clinical trial
investigating a new treatment for rheumatoid arthritis, using an
extended mosaic plot with residual-based shading based on the
maximum statistic: clearly, the new treatment is effective. The dark
blue cell indicates that the rate of treated patients showing marked improvement
is significant at the 1\% level. Figure
\ref{fig:ucbadmissions} visualizes the well-known UCB admissions data
by means of a conditional association plot. The panels show the
residuals from a conditional independence model (independence of gender
and admission, given department), stratified by department. Clearly,
the situation in department A (more women/less men accepted than
would be expected under the null hypothesis) causes the rejection of the
hypothesis of conditional independence. Figure~\ref{fig:presex}
illustrates the conditional independence of premarital and
extramarital sex, given gender and marital status. The $\chi^2$ test 
of independence, based on the permutation distribution,  
rejects the null hypothesis: possibly, because
the tendency of people to have extramarital
sex when they had premarital sex 
is particularly marked among married people? The rate of such women and men
ist significant at the 0.01 and 0.1 level, respectively.
Finally, Figure~\ref{fig:titanic} visualizes the ``Survival on the Titanic'' 
data using a double-decker plot. Here, a binary response (survival of the disaster)
is to be explained by other factors (class, gender, and age). The gray
boxes represent the proportion of survived passengers in a particular
stratum. The proportions of saved women and children are indeed higher
than those of men, but they clearly decrease from the 1st to the 3rd class. In
addition, the proportion of saved men in the 1st class is higher than
in the others.

@@

\clearpage
\section[Labeling]{Labeling}
\label{sec:labeling}

@@

>>>>

\documentclass[article]{jss}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% declarations for jss.cls %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%% almost as usual
\author{David Meyer\\Wirtschaftsuniversit\"at Wien \And 
        Achim Zeileis\\Wirtschaftsuniversit\"at Wien \And 
        Kurt Hornik\\Wirtschaftsuniversit\"at Wien}
\title{The Strucplot Framework:\\ Visualizing Multi-way Contingency
  Tables with \pkg{vcd}}

%% for pretty printing and a nice hypersummary also set:
\Plainauthor{David Meyer, Achim Zeileis, Kurt Hornik} %% comma-separated
\Shorttitle{The Strucplot Framework} %% a short title (if necessary)

%% an abstract and keywords
\Abstract{
This paper describes the ``strucplot'' framework for the visualization of
multi-way contingency tables. Strucplot displays include hierarchical
conditional plots such as mosaic, association, and sieve plots, and
can be combined into more complex, specialized plots for visualizing
conditional independence, GLMs, and the results of
independence tests. The framework's modular design allows
flexible customization of the plots' graphical appearance, including 
shading, labeling, spacing, and legend, by means of ``graphical
appearance control'' functions. 
The framework is provided by the \proglang{R} package \pkg{vcd}.
}

@@

%\VignetteIndexEntry{The Strucplot Framework: Visualizing Multi-way Contingency Tables With package vcd}
%\VignetteDepends{vcd}
%\VignetteKeywords{contingency tables, mosaic plots, association plots, sieve plots, R, categorical data, independence, conditional independence, HSV, HCL, residual-based shading, grid}
%\VignettePackage{vcd}

%% publication information
%% NOTE: This needs to filled out ONLY IF THE PAPER WAS ACCEPTED.
%% If it was not (yet) accepted, leave them commented.
%% \Volume{13}
%% \Issue{9}
%% \Month{September}
%% \Year{2004}
%% \Submitdate{2004-09-29}
%% \Acceptdate{2004-09-29}

%% The address of (at least) one author should be given
%% in the following format:
\Address{
  David Meyer\\
  Department of Information Systems and Operations\\
  E-mail: \email{David.Meyer@wu-wien.ac.at}\\
  URL: \url{http://wi.wu-wien.ac.at/~meyer/}\\

  Achim Zeileis\\
  Department of Statistics and Mathematics\\
  E-mail: \email{Achim.Zeileis@wu-wien.ac.at}\\
  URL: \url{http://statmath.wu-wien.ac.at/~zeileis/}\\

  Kurt Hornik\\
  Department of Statistics and Mathematics\\
  E-mail: \email{Kurt.Hornik@wu-wien.ac.at}\\
  URL: \url{http://statmath.wu-wien.ac.at/~hornik/}\\

  Wirtschaftsuniversit\"at Wien\\
  Augasse 2--6
  1090 Wien, Austria\\
}
%% It is also possible to add a telephone and fax number
%% before the e-mail in the following format:
%% Telephone: +43/1/31336-5053
%% Fax: +43/1/31336-734

%% for those who use Sweave please include the following line (with % symbols):
%% need no \usepackage{Sweave.sty}

\SweaveOpts{engine=R,eps=TRUE,height=6,width=7,results=hide,fig=FALSE,echo=TRUE}
\setkeys{Gin}{width=0.7\textwidth}

@@

\begin{enumerate}
\item Varying the shape of bar plots and mosaic displays to yield, e.g.,
  double-decker plots \citep{vcd:hofmann:2001}, spine plots, or
  spinograms \citep{vcd:hofmann+theus}.
\item Using residual-based shadings to visualize log-linear models
  \citep{vcd:Friendly:1994,vcd:Friendly:2000} and significance of
  statistical tests \citep{vcd:Meyer+Zeileis+Hornik:2003,vcd:Zeileis+Meyer+Hornik:2005}.
\item Using pairs plots and trellis-like layouts for marginal, conditional and
  partial views \citep{vcd:Friendly:1999}.
\item Adding direct user interaction, allowing quick exploration and
  modification of the visualized models 
  \citep{vcd:Unwin+Hawkins+Hofmann:1996,vcd:Theus:2003}.
\item Providing a modular and flexible implementation to easily allow
  user extensions \citep{vcd:Meyer+Zeileis+Hornik:2003}.
\end{enumerate}

\noindent Current implementations of mosaic displays can be found,
e.g., for \proglang{SAS} \citep{vcd:SAS:2005}, \pkg{ViSta} \citep{vcd:young:1996}, 
\pkg{MANET} \citep{vcd:Unwin+Hawkins+Hofmann:1996},
\pkg{Mondrian} \citep{vcd:Theus:2003},
\proglang{R} \citep{vcd:R:2006}, and \proglang{S-PLUS}
\citep{vcd:SPLUS:2005}. For \proglang{R}, currently three
implementations do exist in the packages \pkg{graphics} (in base \proglang{R}), 
\pkg{vcd} \citep{vcd:Meyer+Zeileis+Hornik:2006},
and \pkg{iplots} \citep{vcd:urbanek+wichtrey:2006}, respectively.
Table \ref{tab:compare} gives an overview of the available functionality in these systems.
Most environments are available on Windows, MacOS, and Linux/Unix
variants, except \pkg{MANET} which is only available for the
Macinthosh platforms.

@@

Figures \ref{fig:arthritis} to \ref{fig:titanic} illustrate some of these
extensions (the code is given in Appendix \ref{sec:code}). 
Figure~\ref{fig:arthritis} shows the results from a double-blind clinical trial
investigating a new treatment for rheumatoid arthritis, using an
extended mosaic plot with residual-based shading based on the
maximum statistic: clearly, the new treatment is effective. The dark
blue cell indicates that the rate of treated patients showing marked improvement
is significant at the 1\% level. Figure
\ref{fig:ucbadmissions} visualizes the well-known UCB admissions data
by means of a conditional association plot. The panels show the
residuals from a conditional independence model (independence of gender
and admission, given department), stratified by department. Clearly,
the situation in department A (more women/less men accepted than
would be expected under the null hypothesis) causes the rejection of the
hypothesis of conditional independence. Figure~\ref{fig:presex}
illustrates the conditional independence of premarital and
extramarital sex, given gender and marital status. The $\chi^2$ test 
of independence, based on the permutation distribution,  
rejects the null hypothesis: possibly, because
the tendency of people to have extramarital
sex when they had premarital sex 
is particularly marked among married people? The rate of such women and men
ist significant at the 0.01 and 0.1 level, respectively.
Finally, Figure~\ref{fig:titanic} visualizes the ``Survival on the Titanic'' 
data using a double-decker plot. Here, a binary response (survival of the disaster)
is to be explained by other factors (class, gender, and age). The gray
boxes represent the proportion of survived passengers in a particular
stratum. The proportions of saved women and children are indeed higher
than those of men, but they clearly decrease from the 1st to the 3rd class. In
addition, the proportion of saved men in the 1st class is higher than
in the others.

@@

\section[Labeling]{Labeling}
\label{sec:labeling}

@@

\section{Code used in the introduction}
\label{sec:code}

\begin{itemize}

\item Mosaic plot for the \data{Arthritis} data in Figure~\ref{fig:arthritis}:
\begin{Sinput}
mosaic(art, gp = shading_max, split_vertical = TRUE)
\end{Sinput}

\item Conditional association plot for the \data{UCBAdmissions} 
data in Figure~\ref{fig:ucbadmissions}:
\begin{Sinput}
cotabplot(UCBAdmissions, panel = cotab_indep, shade = TRUE, legend = FALSE)
\end{Sinput}

\item Mosaic plot for the \data{PreSex} data in Figure~\ref{fig:presex}:
\begin{Sinput}
presextest <- coindep_test(PreSex, margin = c(1,4), 
                           indepfun = function(x) sum(x^2), n = 5000)
mosaic(PreSex, condvars = c(1, 4), shade = TRUE, 
       gp_args = list(p.value = presextest$p.value)) %$
\end{Sinput}

\item Double-decker plot for the \data{Titanic} data in Figure~\ref{fig:titanic}:
\begin{Sinput}
doubledecker(Survived ~ ., data = Titanic, labeling_args = 
             list(set_varnames = c(Sex = "Gender")))
\end{Sinput}
\end{itemize}
